---
title: "Assignment 4"
author: "Lauren Chetley"
format: html
editor: visual
---

## Define the Contract

This function calculates Simpson's Index from a list of species names.

The formula that will be used is: D = sum(n \* (n - 1)) / (N \* (N-1))

-   n is the number of individuals in one species

-   N is the total number of individuals of all species

Inputs and Outputs

-   @param species: vector name of species

-   @return: Simpson's index D

    ## Implement the R function

```{r}
# Simpsons_index created the function with species as the input. We will define species names later. 
# N counts the total number of individuals.
# Simpsons formula requires at least two individuals. Fewer than two individuals will result in an error message. 
simpsons_index=function(species){
N=length(species)
  if(N<2){stop("You need at least two individuals")}

# Empty vector to store species names and counts. 
# These empty vectors will be used for future looping. 
species_names=c() #store unique species names
species_counts=c() #store counts of each species 

#This loop goes through each species one at a time. If the species has been seen before, then it adds 1 to the count.The loops add a new species to the list with count of 1, if it has not been seen before. 
 for(i in 1:N) {
    current_species=species[i]
    match_count=0 
    if (length(species_names)>0) { 
      for (j in 1:length(species_names)) {
        if(current_species==species_names[j]) {
          species_counts[j]=species_counts[j] + 1 
          match_count=match_count + 1
        }
      }
    }
    if (match_count==0) {
      species_names=c(species_names, current_species)
      species_counts=c(species_counts, 1)
    }
 }

#Compute numerator of Simpson's formula: sum(n * (n - 1)).
#This loops through each species count, computing (n * (n - 1)) for that species, and then adds it to a running total. 
  numerator=0 
  for(i in 1:length(species_counts)) {
    n=species_counts[i]
    numerator=numerator + n * (n - 1)
  }

#Commpute Simpson's Index
result=numerator  / (N * (N - 1))
  return(result)
}
```

## Testing the Function with Sample Data

```{r}
# Example one: three species with the same number of individuals 
species_ex_1 = c("alligator", "panda", "penguin", "alligator", "panda", "penguin", "alligator", "panda", "penguin" )
D_ex_1 = simpsons_index(species_ex_1)
print(D_ex_1)

#Output: [1] 0.25

# Example two: one species is more common than the others
species_ex_2 = c("panda", "panda", "panda", "panda", "panda", "penguin", "alligator")
D_ex_2 = simpsons_index(species_ex_2)
print(D_ex_2)

#Output: [1] 0.4761905

# Example three: all individuals are the same 
species_ex_3 = c("panda", "panda", "panda", "panda", "panda", "panda", "panda")
D_ex_3 = simpsons_index(species_ex_3)
print(D_ex_3)

#Output: [1] 1

```
